\chapter*{Acknowledgments}

Undertaking this PhD ...

First and foremost, I would like to thank my supervisor Prof. Dr. Andreas Dreuw for taking me under his wing for the last three years. He created a research environment that allowed for a lot of freedom on how to approach the project, and he was there for the moments where I felt lost during it.

A big thanks goes to all members of his research group as well. I had many serious (and not so serious) discussions with them that greatly helped my progress. Special thanks to Maximilian Scheurer for his input on programming practices and being one of the proof readers for this thesis.

I would also like to thank Prof. Dr. Christian Ochsenfeld and his group for giving me insight on the concepts of linear scaling algorithms during the time I could spend in Munich.

My thanks also go out to Prof. Dr. Patrick Norman from KTH for allowing me access to the Tegner and Beskow supercomputer clusters at the PDC in Stockholm. Without these computing resources, the development of \mchem{} would not have progressed as smoothly and rapidly as it did. 

I also like to thank the research group at NTNU: Prof. Dr. Henrik Koch for his online-course on electronic structure theory, and Alexander Paul for being the second proof-reader for this thesis. 

I am also indebted to Prof. Dr. Sonia Coriani, who made me aware of this position in the first place.   

I gratefully acknowledge the funding received from the Marie Skłodowska-Curie European Training Network "COSINE - COmputational Spectroscopy In Natural sciences and Engineering” (Grant agreement no. 765739). It is regretful that the COVID-19 pandemic had such a profound impact on the international training events.

Last but not least, I would like to thank my parents for their support all throughout my Bachelor, Master and PhD studies. Without them, I would not be where I am now. 